Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016

### 2019-08-30 anzjs_elections.Rmd
updated spatial coefficient section - removed electorates voting independently. minor edit to robustness. shift around effects in Influential Socio-demographics. edited conclusion.

The spatial autoregressive coefficient $\rho$ is positive and significant in only the 2001 and 2016 elections (Figure \@ref(fig:rhovis)), meaning that in these elections, the political climate of an electorate appears to be affected by the attitudes of its neighbours. Conversely, in the other four elections, the spatial effect weakens to become insignificant. In these years, it appears that the spatial component does not explain anything not already explained by the electoral socio-demographics, meaning electorates effectively voted independently.

@@

From 2007 onwards, both marriages (`Married`) and de facto relationships (`DeFacto`) are found to be strong predictors of the two-party preferred vote in opposition to the Liberal party. In 2001 and 2004 neither of these variables were significant (see rows 4 and 5 in Figure \@ref(fig:partresplot2)).

@@

@@

Two robustness checks are conducted to confirm model stability. First, a model for each election is fit using only the variables that were found to be significant in at least one of the six elections. The estimated coefficients of the variables in these models all fall within their respective $95\%$ confidence intervals from the full models, confirming that the significant effects are robust to variable selection. The second check involves computing pairwise correlations amongst the explanatory variables, and seeing if the removal of one variable (from a pair) affects the other's estimated effect. For example, the variables `Born_Asia` and `BornElsewhere` have a pairwise correlation of $0.93$ - the largest of any pair. A model is fit omitting `BornElsewhere` (the reduced model). The estimated coefficent of `Born_Asia` is then compared with the confidence interval for `Born_Asia` in the full model. It is found that for each of the top ten pairwise correlations, the estimated effect in the reduced model lies within the confidence interval from the full model. This second check indicates that variables with high pairwise correlation are not masking the effects of each other.

@@

Across the past six elections, most of the socio-demographics that drive the electoral two-party preferred vote are found to remain steady, whilst a few (typically weaker) effects vary over time. Industry and type of work are particularly influential, with energy-related and manufacturing/construction jobs, as well as administrative roles being strongly linked with the Liberal party in all elections. Incomes have a similarly consistent effect, with higher income areas supporting Liberal. Higher levels of unemployment shift from weak association with Labor to a significant Liberal effect over the years, and higher education levels are associated with Labor from 2007 (although significant only in 2016). It is also found that electorates with higher household mobility support Liberal, birthplace diversity favours Labor and more de facto relationships align with Labor preference --- although marriages, family and household sizes have no material influence. Furthermore, the neighbourhood (spatial) effects are found to be positive in all elections, although significant only in 2001 and 2016, meaning that in the 2004--2013 elections, electorates effectively voted independently.

>>>>

The spatial autoregressive coefficient $\rho$ is positive and significant in the 2001, 2004 and 2016 elections (Figure \@ref(fig:rhovis)). In these three elections, there is evidence to suggest that the political climate of an electorate appears to be affected by the attitudes of its neighbours, affecting the electorate's two-party preference. Conversely, in the other three elections, the spatial effect weakens to become insignificant (although still positive).

@@

@@

From 2007 onwards, both marriages (`Married`) and de facto relationships (`DeFacto`) are found to be strong predictors of the two-party preferred vote in opposition to the Liberal party. In 2001 and 2004 neither of these variables are significant (see rows 4 and 5 in Figure \@ref(fig:partresplot2)).

@@

Two robustness checks are conducted to confirm model stability. First, a model for each election is fit using only the variables that are found to be significant in at least one of the six elections. The estimated coefficients of the variables in these models all fall within their respective $95\%$ confidence intervals from the full models, confirming that the significant effects are robust to variable selection. The second check involves computing pairwise correlations amongst the explanatory variables, and seeing if the removal of one variable (from a pair) affects the other's estimated effect. For example, the variables `Born_Asia` and `BornElsewhere` have a pairwise correlation of $0.93$ - the largest of any pair. A model is fit omitting `BornElsewhere` (the reduced model). The estimated coefficent of `Born_Asia` is then compared with the confidence interval for `Born_Asia` in the full model. It is found that for each of the top ten pairwise correlations, the estimated effect in the reduced model lies within the confidence interval from the full model. This second check indicates that variables with the highest pairwise correlation are not masking the effects of each other.

@@

Across the past six elections, most of the socio-demographics that drive the electoral two-party preferred vote are found to remain steady, whilst a few (typically weaker) effects vary over time. Industry and type of work are particularly influential. Energy-related and manufacturing/construction jobs, as well as administrative roles and jobs in education and social services are strongly linked with the Liberal party in all elections. Incomes have a similarly consistent effect, with higher income areas supporting Liberal. Higher levels of unemployment shift from weak association with Labor to a significant Liberal effect over the years, and higher education levels are associated with Labor from 2007 (although marginally insignificant in 2010). Electorates with large populations 20 to 34 years are strongly associated with Labor, whilst the 35 to 54 year old bracket also increases Labor support, but to a lesser extent. It is also found that birthplace diversity slightly favours Labor, relationships (both marriages and de facto relationships) align with Labor preference from 2010 onwards, and the influence of Christian populations has trended towards Liberal support whilst other religions have negligible effects. Family and household sizes have minimal influence, although electorates with more single parent households are linked with Labor support. Furthermore, the neighbourhood (spatial) effects are found to be positive in all elections and significant in 2001, 2004 and 2016, meaning that neighbouring party preference appears to be an influential factor in those years.
